---
title: 'EDS 223: assignment 2'
author: "Charlie Curtin"
output: html_document
date: "2023-10-17"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Background

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930's the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York
Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.[^4] Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Check out coverage by
[EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

### Data

#### EJScreen

We will be working with data from the United States Environmental
Protection Agency's EJScreen: Environmental Justice Screening and
Mapping Tool.

According to the [US EPA
website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents
> or other stakeholders as they search for environmental or demographic
> information. It can also support a wide range of research and policy
> goals. The public has used EJScreen in many different locations and in
> many different ways.
>
> EPA is sharing EJScreen with the public:\
> - to be more transparent about how we consider environmental justice
> in our work,\
> - to assist our stakeholders in making informed decisions about
> pursuing environmental justice and,\
> - to create a common starting point between the agency and the public
> when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels.
We will be working with block group data that has been downloaded from
the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data).

#### Mapping Inequality

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

We will be working with maps of HOLC grade designations for Los Angeles.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

#### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed.

We will be working with observations of birds from 2021 onward.

## Data Analysis

To frame inequity in Los Angeles, we'll first map percentile proximity to wastewater discharge by census block group. We'll overlay the centroids of census block groups with a percentile proximity to wastewater discharge higher than 95.

```{r}
# load packages
library(tidyverse)
library(sf)
library(tmap)
library(knitr)
library(kableExtra)
```

```{r}
# read in the EJScreen data
ejscreen <- st_read('data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb',
                    quiet = TRUE)

# filter to Los Angeles County
los_angeles <- ejscreen %>% 
  filter(CNTY_NAME %in% c("Los Angeles County"))
```

We'll create a new layer with the centroids of each census block groups and filter for block groups with a wastewater proximity percentile higher than 95. We'll create a chloropleth map, with census block groups colored by percentile proximity to wastewater, and the filtered centroids represented as red points.

```{r }
# find the centroid within each block group to plot the block groups as points
LA_centroids <- st_centroid(los_angeles)

# filter to centroids where percentile of wastewater discharge is above 95
LA_centroids_percentile <- LA_centroids %>% 
  filter(P_PWDIS > 95)

# map the percentile proximity to wastewater discharge by census block groups and add centroids for block groups above the 95th percentile
map1 <- tm_shape(los_angeles) +
  tm_fill(col = 'P_PWDIS',
          palette = 'Blues',
          n = 6,
          title = 'percentile PWDIS') +
  tm_layout(main.title = "Percentile Proximity to Wastewater Discharge \n by Census Block Groups in LA County",
            main.title.size = 1,
            legend.outside = TRUE,
            legend.frame = TRUE) +
  tm_compass(position = c('right', 'bottom')) +
  tm_scale_bar(position = c('right', 'bottom')) +
  tm_add_legend(type = 'symbol',
                title = 'block group centroids',
                labels = c('above 95th percentile'),
                col = 'red')

map1 +
  tm_shape(LA_centroids_percentile) +
  tm_dots(col = 'red')
```

Let's find the percentage of census block groups where less than 5% of the population is considered low-income and block groups above the 80th percentile for PM 2.5 and Superfund proximity.

```{r }
# find the percent of census block groups where less than 5% of the population is considered low-income 
print(paste("Percentage of census block groups in LA County where less than 5% of the population is considered low-income:", round(nrow(filter(los_angeles, LOWINCPCT < .05)) / nrow(los_angeles) * 100, 2)))

# find the percent of block groups that are above the 80th percentile for PM 2.5 and Superfund Proximity
print(paste("Percentage of census block groups in LA County that are above the 80th percentile for PM 2.5 and Superfund proximity:", round(nrow(filter(los_angeles, (P_PM25 > 80 & P_PNPL > 80))) / nrow(los_angeles)* 100, 2)))
```

Let's plot historical redlining boundaries colored by HOLC grade, with census block groups for context. First, we'll transform our redlining boundaries to match the coordinate reference system to the census block groups before mapping.

```{r }
# read in the redlining data
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json", quiet = TRUE) %>%
  st_make_valid()

# match the CRS
LA_redlining <- st_transform(LA_redlining, st_crs(los_angeles))

# map historical redlining boundaries colored by HOLC grade using LA block groups for context
base <- tm_shape(los_angeles, bbox = st_bbox(LA_redlining)) +
  tm_borders(lwd = .3,
             alpha = .5)

base +
  tm_shape(LA_redlining) +
  tm_polygons(col = 'grade',
              palette = 'viridis',
              title = 'HOLC grade') +
  tm_layout(main.title = 'Historical Redlining Grades in Los Angeles, CA',
            main.title.size = 1,
            legend.frame = TRUE,
            inner.margins = c(0.1, 0.1, 0.1, 0.1)) +
  tm_compass(position = c('left', 'bottom')) +
  tm_scale_bar(position = c('left', 'bottom'))
```


Next, we'll find the number of census block groups that fall within historical redlining boundaries. To do this, we'll use spatial subsetting, which will create a new layer of all census block groups that intersect with the polygons in the redlining data.

```{r}
# use spatial subsetting to find block groups that fall within the boundaries of historical redlining data
bg_redline <- los_angeles[LA_redlining, ]

print(paste("There are", nrow(bg_redline), "census block groups that fall within the boundaries of historical redlining"))
```

To summarize current conditions in redlined areas, let's find the mean of percent low income, percentile PM 2.5, percentile low life expectancy, and percentile air toxics cancer risk by historical redlining category.

```{r }
# join our layers to get the historical redlining grade in each census block group
LA <- st_intersection(LA_redlining, los_angeles)

# find mean of % low income, percentile PM 2.5, percentile low life expectancy, and percentile air toxics cancer risk by historical redlining categories
redline_summary <- join %>% 
  group_by(grade) %>% 
  summarize(linc_mean = (mean(LOWINCPCT, na.rm = TRUE) * 100),
            pm25_mean = mean(P_PM25, na.rm = TRUE),
            plife_mean = mean(P_LIFEEXPPCT, na.rm = TRUE),
            pcancer_mean = mean(P_CANCER, na.rm = TRUE)) %>% 
  st_drop_geometry()

# view the dataframe
redline_summary %>% 
  kable("html") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

For each variable, we can see that the mean increases as the HOLC grade moves from A to D. Banks used the HOLC grades to inform their lending practices, making it more difficult for residents to finance homes as the grades moves from A to D. Our calculations provide evidence that the legacy of redlining still clearly shapes economic and health disparity between census block groups today.

### Investigate the legacy of redlining in biodiversity observations

Let's find the percentage of bird observations from 2022 that fall within each redlining category. For this part of the analysis, we'll need to find the number of bird observations in each HOLC grade.

```{r}
# read in the bird observations data
birds <- st_read('data/gbif-birds-LA', quiet = TRUE)

# filter to observations in 2022
birds_22 <- birds %>% 
  filter(year == 2022)

# change the CRS of birds to match bg_redline
birds_22 <- st_transform(birds_22, st_crs(LA_redlining))

# count the number of bird observations for areas with HOLC grades and add it as a column
LA_redlining$bird_obs <- lengths(st_intersects(LA_redlining, birds_22))

# find the percent of observations within each redlining category
obs_total <- sum(LA_redlining$bird_obs)

birds_summ <- LA_redlining %>%
  group_by(grade) %>% 
  summarize(p_obs = (sum(bird_obs) / obs_total) * 100,
            total_obs = sum(bird_obs)) %>% 
  st_drop_geometry()

# view the summary table  
birds_summ %>% 
  kable("html") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

Now let's plot our results as a bar chart.
```{r}
# plot the results
birds_summ %>% 
  ggplot(aes(x = grade, y = p_obs)) +
  geom_bar(color = 'black',
           fill = 'seagreen',
           stat = 'identity') +
  theme_bw() +
  labs(title = 'Percent of Bird Observations by HOLC Grade in Los Angeles',
       x = 'HOLC grade',
       y = 'percent')
  
```

These results are surprising, as they don't align with Ellis-Soto's findings that redlined neighborhoods are more undersample than other areas. Areas with a HOLC grade of C and D represent the highest percentage of bird observations. Due to disparities in greenery between redlined areas, the acutal biodiversity might be lower in these areas. One possible explanation is that in 2022, there was a citizen science initiative that specifically targeted redlined areas, leading to more observations there.